---
title: "Simulation Vignette: Learning Sensitivity and the ob < ow Pattern"
subtitle: "A transparent account of the code, the mathematics, and the linking hypotheses"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

------------------------------------------------------------------------

# Purpose of this document

This vignette is written for a reader who wants to evaluate three
things:

1.  **Does the code do what it says it does?** Each implementation step
    is linked back to the mathematical formula it is meant to compute,
    with a worked numerical example the reader can verify by hand.

2.  **Is the linking hypothesis valid?** Each model is presented with a
    clear statement of the theoretical claim it is meant to formalise,
    and an argument for why the code implements that claim rather than
    something else.

3.  **Are the conclusions warranted?** The results of each simulation
    are stated as predictions before the code is run, so the reader can
    evaluate whether the claimed findings follow from the setup.

The document covers three models that share identical data generation
but differ in how the learner estimates verb distributions. The shared
components are described once in §2. Each model is then described in
§3–5 with its own theoretical motivation, mathematical formulation, and
worked example.

------------------------------------------------------------------------

# Shared components: data generation and evaluation

All three models use the same procedure to (a) build verb token pools,
(b) construct true verb distributions, and (c) evaluate onset criteria.
This section describes each component in full.

## The theoretical background: what are we simulating?

Jian & Manning (2026) track how GPT-2's next-token distributions change
over training for two classes of verbs: *to-Dative* verbs (e.g. *give*,
*sell*, *award*; N = 35) and *Motion* verbs (e.g. *go*, *run*, *walk*; N
= 36). They measure two onset events:

-   **ob (between-class onset)**: the training step at which class-level
    structure first becomes statistically detectable — the model treats
    dative and motion verbs as systematically different.
-   **ow (within-class onset)**: the training step at which within-class
    idiosyncratic differentiation first crosses an absolute divergence
    threshold — individual verbs start looking distinct from one another
    even within the same class.

J&M find **ob \< ow**: class structure is detectable before item-level
idiosyncrasy reaches threshold. They interpret this as evidence of an
abstraction-first learning bias in GPT-2.

**Our counter-argument**: ob \< ow is a structural consequence of *low
sensitivity to individual observations*, not of abstraction-first
learning. Any learner that requires many consistent observations before
its estimate shifts will detect the stronger class-level signal before
the noisier item-level signal. We demonstrate this with three
simulations.

## Token pool construction

### The theoretical motivation

Each verb in the simulation has a set of "preferred" tokens — tokens
that appear with elevated probability in that verb's context. We
partition these into three groups:

| Group | Shared with | Linguistic analogue |
|-----------------|---------------------|----------------------------------|
| **cross tokens** | All verbs in both classes | High-frequency function words, common syntactic frames |
| **within tokens** | All verbs in the same class only | Class-specific argument preferences (e.g. dative verbs near recipient NPs) |
| **idiosyncratic tokens** | Only this verb | Verb-specific collocates (*give* with *permission*; *run* with *marathon*) |

The key structural constraint is **item_overlap \> class_overlap**:
within-class sharing exceeds cross-class sharing. This is what makes
class membership a real distributional distinction. It is also the
critical assumption our argument rests on — if this did not hold, there
would be no class signal to detect.

### The mathematics

Given parameters `class_overlap`, `item_overlap`, and `n_preferred`:

$$n_\text{cross} = \text{round}(\texttt{class\_overlap} \times n_\text{preferred})$$
$$n_\text{within} = \text{round}((\texttt{item\_overlap} - \texttt{class\_overlap}) \times n_\text{preferred})$$
$$n_\text{idio} = n_\text{preferred} - n_\text{cross} - n_\text{within}$$

Token IDs are assigned contiguously. Cross tokens occupy IDs 1 through
$n_\text{cross}$. Within-class tokens for class A occupy the next
$n_\text{within}$ IDs, and similarly for class B. Each verb
independently samples $n_\text{idio}$ idiosyncratic token IDs from the
remaining vocabulary.

### Worked example

With `n_preferred = 10`, `class_overlap = 0.3`, `item_overlap = 0.6`,
`VOCAB_SIZE = 20`:

```{r token_pool_example}
# Parameters
n_preferred   <- 10
class_overlap <- 0.3
item_overlap  <- 0.6
VOCAB_SIZE    <- 20

# Compute group sizes
n_cross  <- round(class_overlap * n_preferred)   # 3 tokens
n_within <- round((item_overlap - class_overlap) * n_preferred)  # 3 tokens
n_idio   <- n_preferred - n_cross - n_within      # 4 tokens

cat("Cross tokens:", n_cross, "\n")
cat("Within-class tokens per class:", n_within, "\n")
cat("Idiosyncratic tokens per verb:", n_idio, "\n")

# Token ID ranges
cross_tok    <- seq_len(n_cross)                            # IDs 1:3
within_A_tok <- n_cross + seq_len(n_within)                 # IDs 4:6
within_B_tok <- n_cross + n_within + seq_len(n_within)      # IDs 7:9
idio_pool    <- (n_cross + 2 * n_within + 1):VOCAB_SIZE     # IDs 10:20

cat("\nCross token IDs:", cross_tok, "\n")
cat("Within-A token IDs:", within_A_tok, "\n")
cat("Within-B token IDs:", within_B_tok, "\n")
cat("Idiosyncratic pool IDs:", head(idio_pool, 5), "...\n")

# Build two example class-A verbs
set.seed(1)
verb_A1 <- c(cross_tok, within_A_tok, sample(idio_pool, n_idio, replace = FALSE))
verb_A2 <- c(cross_tok, within_A_tok, sample(idio_pool, n_idio, replace = FALSE))
# Build one class-B verb
verb_B1 <- c(cross_tok, within_B_tok, sample(idio_pool, n_idio, replace = FALSE))

cat("\nVerb A1 preferred tokens:", sort(verb_A1), "\n")
cat("Verb A2 preferred tokens:", sort(verb_A2), "\n")
cat("Verb B1 preferred tokens:", sort(verb_B1), "\n")

# Check: A1 and A2 share cross + within-A tokens; B1 shares only cross tokens with A1
shared_A1_A2 <- intersect(verb_A1, verb_A2)
shared_A1_B1 <- intersect(verb_A1, verb_B1)
cat("\nTokens shared between A1 and A2:", length(shared_A1_A2), "(cross + within-A)\n")
cat("Tokens shared between A1 and B1:", length(shared_A1_B1), "(cross only)\n")
cat("item_overlap > class_overlap?", length(shared_A1_A2)/n_preferred, ">",
    length(shared_A1_B1)/n_preferred, "\n")
```

The final check confirms the structural asymmetry: same-class verb pairs
share more tokens than cross-class pairs. This is the distributional
signature of class membership that ob is designed to detect.

## True distribution construction

### The theoretical motivation

Each verb's "true" distribution $P_\text{verb}$ is what the learner
would converge to given unlimited data. Preferred tokens receive high
probability; all other tokens receive a low uniform background weight.
The lognormal weighting creates heterogeneity within the preferred set —
preferred tokens are not all equally preferred, just as in real
language.

### The mathematics

Each verb's raw weights are:

$$w_t = \begin{cases} \text{LogNormal}(\mu_\text{log}, \sigma) & \text{if } t \in \text{preferred}(v) \\ 1.0 & \text{otherwise} \end{cases}$$

where $\mu_\text{log} = \log(\mu) - \frac{1}{2}\sigma^2$ ensures
$\mathbb{E}[w_t] = \mu$ for preferred tokens. The distribution is then
normalised: $P_\text{verb}(t) = w_t / \sum_{t'} w_{t'}$.

### Worked example

```{r true_dist_example}
BG_WEIGHT <- 1.0
mu    <- 10
sigma <- 1.0

# lognormal location parameter: ensures E[w] = mu
log_mean <- log(mu) - 0.5 * sigma^2
cat("Expected preferred token weight:", exp(log_mean + 0.5 * sigma^2),
    "(should equal mu =", mu, ")\n")

# Build the true distribution for verb_A1 (defined above)
set.seed(42)
w <- rep(BG_WEIGHT, VOCAB_SIZE)
w[verb_A1] <- rlnorm(n_preferred, log_mean, sigma)  # lognormal weights for preferred tokens
P_A1 <- w / sum(w)

cat("\nTotal probability on preferred tokens:", round(sum(P_A1[verb_A1]), 3), "\n")
cat("Total probability on background tokens:", round(sum(P_A1[-verb_A1]), 3), "\n")
cat("(Preferred tokens are", round(sum(P_A1[verb_A1]) / sum(P_A1[-verb_A1]), 1),
    "x more probable in aggregate)\n")
```

**Linking hypothesis check**: preferred tokens are substantially more
probable than background tokens, but the lognormal variation means some
preferred tokens are much more prominent than others — creating
verb-specific distributional "signatures."

## Jensen-Shannon Divergence

### The theoretical motivation

We need a measure of similarity between two verb distributions.
Jensen-Shannon Divergence (DJS) is used because (a) it is symmetric, (b)
it is bounded in [0, log(2)], and (c) it is zero if and only if the two
distributions are identical. This makes it a natural measure of how
"learnable" the distinction between two verbs is.

### The mathematics

For two distributions $P$ and $Q$:

$$\text{DJS}(P, Q) = H\!\left(\frac{P + Q}{2}\right) - \frac{1}{2}\bigl[H(P) + H(Q)\bigr]$$

where $H(P) = -\sum_t P(t) \log P(t)$ is Shannon entropy. Intuitively,
DJS measures how much more uncertain you are about a randomly-drawn
token if you don't know which distribution it came from, compared to if
you do know. It is zero when $P = Q$ and maximised when $P$ and $Q$ have
no overlap.

### Worked example: manual vs code verification

```{r jsd_example}
# Manual calculation for two simple 4-token distributions
P <- c(0.5, 0.3, 0.1, 0.1)
Q <- c(0.1, 0.1, 0.3, 0.5)

H <- function(p) -sum(p * log(p + 1e-30))  # Shannon entropy

M        <- 0.5 * (P + Q)                    # mixture distribution
jsd_manual <- H(M) - 0.5 * (H(P) + H(Q))   # DJS formula
cat("DJS(P, Q) calculated manually:", round(jsd_manual, 6), "\n")
cat("H(P):", round(H(P), 4), "  H(Q):", round(H(Q), 4),
    "  H(mixture):", round(H(M), 4), "\n")

# Sanity checks
cat("\nDJS(P, P) =", round(H(0.5*(P+P)) - 0.5*(H(P)+H(P)), 6),
    "(should be 0 — identical distributions)\n")
cat("Max possible DJS = log(2) =", round(log(2), 4), "\n")
cat("DJS(P, Q) / log(2) =", round(jsd_manual / log(2), 4),
    "(fraction of maximum)\n")
```

The `pairwise_jsd()` function in the model scripts computes this for all
71 × 71 verb pairs simultaneously using matrix operations rather than a
double loop — this is purely a computational optimisation and does not
change the values computed:

```{r pairwise_jsd_function}
# Reproduction of the pairwise_jsd() function used in all three model scripts
VOCAB_SIZE_FULL <- 1000L  # as used in the actual scripts

pairwise_jsd <- function(P) {
  N   <- nrow(P)
  eps <- 1e-30
  # Step 1: per-row entropy H(P_i) for each verb i
  H_vec <- -rowSums(P * log(P + eps))

  jsd_mat <- matrix(0.0, N, N)
  for (i in seq_len(N - 1L)) {
    j_idx <- (i + 1L):N
    # Step 2: mixture M_ij = 0.5*(P_i + P_j) for all j > i simultaneously
    M     <- 0.5 * (matrix(P[i, ], nrow = length(j_idx),
                            ncol = ncol(P), byrow = TRUE)
                    + P[j_idx, , drop = FALSE])
    # Step 3: H(M_ij) for all j > i
    H_M   <- -rowSums(M * log(M + eps))
    # Step 4: DJS = H(mixture) - 0.5*(H_i + H_j), clamped to >= 0
    d     <- pmax(0.0, H_M - 0.5 * (H_vec[i] + H_vec[j_idx]))
    jsd_mat[i, j_idx] <- d
    jsd_mat[j_idx, i] <- d  # DJS is symmetric
  }
  jsd_mat
}

# Verify against manual calculation: wrap P and Q in a 2-row matrix
P_mat <- rbind(P, Q)
jsd_from_function <- pairwise_jsd(P_mat)[1, 2]
cat("DJS from pairwise_jsd():", round(jsd_from_function, 6), "\n")
cat("DJS from manual calculation:", round(jsd_manual, 6), "\n")
cat("Match:", isTRUE(all.equal(jsd_from_function, jsd_manual)), "\n")
```

## Onset criteria

### ob: between-class onset

**Theoretical motivation**: ob should fire when the model has learned to
treat the two verb classes as systematically different — not just when
two specific verbs happen to diverge, but when the between \> within
pattern holds reliably across verbs. A non-parametric test
(Mann-Whitney) is used because it does not assume normality and tests
the *distributional* ordering rather than just the mean.

**What ob actually tests**: for each verb $v$ in class A, we compare two
sets of DJS distances: - **within**: DJS distances from $v$ to all other
class-A verbs - **between**: DJS distances from $v$ to all class-B verbs

A one-tailed Mann-Whitney test asks: are between-class distances
stochastically greater than within-class distances? ob fires when ≥ 10%
of all verbs pass this test (p \< 0.001).

**Why 10%?** J&M describe the ob onset as the "first departure from 0" —
roughly the first step at which *any* verb passes. Our 10% threshold is
more conservative, meaning ob fires *later* than under J&M's criterion.
This makes the ob \< ow result stronger, not weaker.

**Critical property — ob is a RELATIVE test**: it asks whether between
\> within, not whether either is large in absolute terms. This means ob
can fire even when all DJS values are tiny, as long as the structural
between \> within asymmetry is present. This is key to why ob fires
before ow under high-k/low-sensitivity conditions.

```{r ob_criterion_example}
# Worked example: construct a simple case where ob should fire
# 2 class-A verbs, 2 class-B verbs, all with known DJS values

# Construct the DJS matrix manually for illustration:
# DJS(A1,A2) = 0.02 (small — same class)
# DJS(A1,B1) = 0.05 (larger — different class)
# DJS(A1,B2) = 0.06

jsd_example <- matrix(c(
  0.00, 0.02, 0.05, 0.06,   # A1 row
  0.02, 0.00, 0.04, 0.05,   # A2 row
  0.05, 0.04, 0.00, 0.02,   # B1 row
  0.06, 0.05, 0.02, 0.00    # B2 row
), nrow = 4, byrow = TRUE)

N_A_ex <- 2; N_B_ex <- 2

# ob check for verb A1 (row 1)
within_A1  <- jsd_example[1, setdiff(seq_len(N_A_ex), 1)]    # DJS to A2
between_A1 <- jsd_example[1, (N_A_ex + 1):(N_A_ex + N_B_ex)] # DJS to B1, B2

cat("A1 within-class distances:", within_A1, "\n")
cat("A1 between-class distances:", between_A1, "\n")
test_A1 <- wilcox.test(between_A1, within_A1, alternative = "greater", exact = FALSE)
cat("Mann-Whitney p-value for A1:", round(test_A1$p.value, 4), "\n")
cat("A1 passes (p < 0.001)?", test_A1$p.value < 0.001, "\n")
# Note: with only 2 values in each group, exact p-values are not informative —
# this example illustrates the structure, not statistical power
```

### ow: within-class onset

**Theoretical motivation**: ow fires when within-class idiosyncrasy is
detectable — individual verbs within the same class have started to
diverge from one another. This is measured as the mean DJS across all
same-class verb pairs.

**What ow actually tests**: the mean of the upper triangle of the
within-class DJS sub-matrices (upper triangle to avoid counting
zero-diagonal entries and double-counting pairs).

**Why the threshold of 0.01?** This is taken directly from J&M's
Experiment 2 breakpoint criterion: within-class DJS is "consistently at
least 0.01 greater than baseline."

**Why 3 consecutive steps?** The sustained-window requirement prevents a
single noisy spike from triggering ow prematurely, matching J&M's
requirement that the threshold be exceeded "consistently."

**Critical property — ow is an ABSOLUTE test**: it asks whether
within-class DJS reaches 0.01, regardless of what between-class DJS is
doing. This is what makes the ob/ow ordering sensitive to learning
sensitivity (k or γ): under low sensitivity, both ob (relative) and ow
(absolute) start from near-zero DJS, and ob's relative test detects the
structural asymmetry first; under high sensitivity, ow's threshold is
crossed by early idiosyncratic spiking before ob's relative test becomes
reliable.

```{r ow_criterion_example}
# Worked example of the ow streak requirement
# Simulate a sequence of mean within-class DJS values over training steps
DJS_THRESH <- 0.01
SUSTAIN    <- 3L

within_djs_trajectory <- c(0.002, 0.005, 0.008, 0.012, 0.015, 0.011, 0.009, 0.013, 0.014, 0.016)

ow_nobs   <- NA
ow_streak <- 0L
ow_start  <- NA

for (step in seq_along(within_djs_trajectory)) {
  val <- within_djs_trajectory[step]
  if (val > DJS_THRESH) {
    if (is.na(ow_start)) ow_start <- step
    ow_streak <- ow_streak + 1L
    if (ow_streak >= SUSTAIN) { ow_nobs <- ow_start; break }
  } else {
    ow_streak <- 0L
    ow_start  <- NA
  }
  cat(sprintf("Step %2d: DJS = %.3f  streak = %d  ow_start = %s\n",
              step, val, ow_streak, ifelse(is.na(ow_start), "NA", ow_start)))
}
cat(sprintf("\now fires at step: %s (the first step of the sustained run)\n",
            ifelse(is.na(ow_nobs), "NA (not detected)", ow_nobs)))
```

Note that ow fires at the step where the streak *began*, not the step
where it completed. This is intentional: we want to record when the
sustained period of above-threshold DJS started, not when we accumulated
enough evidence to declare it.

------------------------------------------------------------------------

# Model 1: Zero-sensitivity limit (deterministic interpolation)

## Theoretical claim

**Claim**: if a learner has zero sensitivity to individual observations
— if every new token observation moves the learner's estimate by an
infinitesimal amount — then the learner will always show ob \< ow.

This is the theoretical extreme case. It serves two purposes: (1) it
shows the predicted result under the strongest version of the
low-sensitivity hypothesis, and (2) it provides a clean analytical
baseline that the more realistic models (2 and 3) should converge to as
sensitivity decreases.

## Mathematical formulation

At learning progress $\alpha \in [0, 1]$, each verb's estimated
distribution is:

$$\hat{P}_\text{verb}(\alpha) = (1 - \alpha) \cdot \text{Uniform} + \alpha \cdot P_\text{verb} \tag{1}$$

-   At $\alpha = 0$: all verbs look uniform — the random-initialisation
    state
-   At $\alpha = 1$: each verb shows its true distributional profile
-   At intermediate $\alpha$: a convex mixture

**Why does equation (1) correspond to zero sensitivity?**

A learner with sensitivity controlled by $k$ (Model 2) has expected
distribution:

$$\mathbb{E}[\hat{P}_\text{verb}(n)] = \frac{n}{n + kV} \cdot P_\text{verb} + \frac{kV}{n + kV} \cdot \text{Uniform} = \alpha(n) \cdot P_\text{verb} + (1 - \alpha(n)) \cdot \text{Uniform}$$

where $\alpha(n) = n/(n + kV)$. This is exactly equation (1), with
$\alpha$ determined by corpus size and $k$. As $k \to \infty$, the
variance of the stochastic estimate around this expectation goes to zero
(each observation contributes $\approx 1/(kV) \to 0$ to the estimate),
and Model 2 collapses to this deterministic trajectory. **Model 1 is
the** $k \to \infty$ limit of Model 2.

```{r model1_kinfty_proof}
# Demonstrate numerically: at large k, Model 2 expectation matches Model 1

# True distribution (simplified 5-token example)
P_true <- c(0.40, 0.25, 0.15, 0.12, 0.08)  # sums to 1
V      <- length(P_true)

# At n_obs = 100 with various k values, compute E[P_hat]
n_obs <- 100
k_values <- c(0.001, 0.1, 1, 10, 100)

for (k in k_values) {
  alpha_n <- n_obs / (n_obs + k * V)
  E_Phat  <- alpha_n * P_true + (1 - alpha_n) * (1/V)
  cat(sprintf("k = %5.3f: alpha(n) = %.4f, E[P_hat(1)] = %.4f  (Model 1 value: %.4f)\n",
              k, alpha_n, E_Phat[1], 0.40 * alpha_n + (1/V) * (1 - alpha_n)))
}

# At k=100, alpha(n) is near 0 for small n -- distribution barely moves from uniform
k_large <- 1000
alpha_large <- n_obs / (n_obs + k_large * V)
cat(sprintf("\nAt k = 1000, alpha(100) = %.6f -- distribution is %.1f%% uniform\n",
            alpha_large, (1 - alpha_large) * 100))
```

## Why ob \< ow is guaranteed in Model 1

**Step 1: Between-class DJS is always larger than within-class DJS (for
any α \> 0)**

At any $\alpha > 0$, each verb's distribution is a mixture of uniform
and its true distribution. The DJS between two verbs is proportional to
$\alpha^2$ times the DJS between their true distributions (for small
$\alpha$). Because item_overlap \> class_overlap, same-class verbs share
more preferred tokens than cross-class verbs, so DJS(same class) \<
DJS(different class) by construction. The relative Mann-Whitney test for
ob detects this structural asymmetry.

**Step 2: The absolute ow threshold must wait for α to grow**

At very small $\alpha$, all DJS values are tiny (proportional to
$\alpha^2$). The ow threshold requires mean within-class DJS to reach
0.01. This requires $\alpha$ to be large enough that
$\alpha^2 \cdot \text{DJS}_\text{true, within} \geq 0.01$. Since the
relative ob test has no minimum magnitude requirement, ob fires at a
much smaller $\alpha$.

**Linking hypothesis**: the gap between ob and ow onset is guaranteed
whenever (a) the true distributions have the within \> cross-class
similarity structure, and (b) the trajectory is monotone (no
idiosyncratic intermediate phase). Condition (a) is encoded in the token
pool. Condition (b) is guaranteed by the deterministic interpolation in
equation (1).

## Implementation walkthrough

```{r model1_implementation, eval=TRUE}
# ---- Minimal self-contained implementation of Model 1 ----
# Using small parameters to keep output readable

VOCAB_SIZE_M1 <- 50L
N_PREFERRED   <- 10L
N_A_M1        <- 5L    # small class sizes for illustration
N_B_M1        <- 5L
BG_WEIGHT_M1  <- 1.0

# Onset thresholds
P_THRESH   <- 0.001
CLASS_FRAC <- 0.10
DJS_THRESH <- 0.01
SUSTAIN    <- 3L

# Alpha grid
ALPHAS_M1 <- exp(seq(log(0.001), log(1.0), length.out = 25L))

set.seed(1)

# Step 1: Token pool
n_cross  <- round(0.3 * N_PREFERRED)   # 3 cross tokens
n_within <- round(0.3 * N_PREFERRED)   # 3 within tokens
n_idio   <- N_PREFERRED - n_cross - n_within  # 4 idio tokens

cross_tok    <- seq_len(n_cross)
within_A_tok <- n_cross + seq_len(n_within)
within_B_tok <- n_cross + n_within + seq_len(n_within)
idio_pool    <- (n_cross + 2L * n_within + 1L):VOCAB_SIZE_M1

mk <- function(cls) c(cross_tok, cls, sample(idio_pool, n_idio, replace = FALSE))
A_tokens <- replicate(N_A_M1, mk(within_A_tok), simplify = FALSE)
B_tokens <- replicate(N_B_M1, mk(within_B_tok), simplify = FALSE)

# Step 2: True distributions
log_mean <- log(30) - 0.5 * 1.0^2  # mu=30, sigma=1.0
bld <- function(tok_list) {
  P <- matrix(BG_WEIGHT_M1, nrow = length(tok_list), ncol = VOCAB_SIZE_M1)
  for (i in seq_along(tok_list)) P[i, tok_list[[i]]] <- rlnorm(N_PREFERRED, log_mean, 1.0)
  P / rowSums(P)
}

P_true <- rbind(bld(A_tokens), bld(B_tokens))  # (N_A + N_B) x V
uni    <- 1.0 / VOCAB_SIZE_M1                  # scalar: uniform probability

# Step 3: Sweep alpha and check onsets
ob_alpha <- NA_real_; ow_alpha <- NA_real_; streak <- 0L; ow_start <- NA_real_

for (alpha in ALPHAS_M1) {
  # Equation (1): interpolated distributions
  P_obs <- (1 - alpha) * uni + alpha * P_true

  # Pairwise DJS
  D <- pairwise_jsd(P_obs)

  # ob check (same logic as model scripts)
  if (is.na(ob_alpha)) {
    n_sig <- sum(sapply(seq_len(N_A_M1), function(i) {
      suppressWarnings(wilcox.test(
        D[i, (N_A_M1+1):(N_A_M1+N_B_M1)],
        D[i, setdiff(seq_len(N_A_M1), i)],
        alternative = "greater", exact = FALSE)$p.value) < P_THRESH
    })) +
    sum(sapply(seq_len(N_B_M1), function(j) {
      i <- N_A_M1 + j
      suppressWarnings(wilcox.test(
        D[i, seq_len(N_A_M1)],
        D[i, N_A_M1 + setdiff(seq_len(N_B_M1), j)],
        alternative = "greater", exact = FALSE)$p.value) < P_THRESH
    }))
    if (n_sig / (N_A_M1 + N_B_M1) >= CLASS_FRAC) ob_alpha <- alpha
  }

  # ow check
  if (is.na(ow_alpha)) {
    jA <- D[seq_len(N_A_M1), seq_len(N_A_M1)]
    jB <- D[(N_A_M1+1):(N_A_M1+N_B_M1), (N_A_M1+1):(N_A_M1+N_B_M1)]
    wm <- (mean(jA[upper.tri(jA)]) + mean(jB[upper.tri(jB)])) / 2
    if (wm > DJS_THRESH) {
      if (is.na(ow_start)) ow_start <- alpha
      streak <- streak + 1L
      if (streak >= SUSTAIN) ow_alpha <- ow_start
    } else { streak <- 0L; ow_start <- NA_real_ }
  }
  if (!is.na(ob_alpha) && !is.na(ow_alpha)) break
}

cat("ob fires at alpha =", round(ob_alpha, 5), "\n")
cat("ow fires at alpha =", round(ow_alpha, 5), "\n")
cat("ob < ow?", ob_alpha < ow_alpha, "\n")
cat("log(ow/ob) =", round(log(ow_alpha / ob_alpha), 3),
    "(positive = ob fired first; larger = bigger gap)\n")
```

**Linking hypothesis confirmed**: ob fires at a much smaller $\alpha$
than ow. The relative criterion (ob) detects the structural between \>
within asymmetry immediately; the absolute criterion (ow) must wait for
DJS to build up.

------------------------------------------------------------------------

# Model 2: Add-k smoothing

## Theoretical claim

**Claim**: the ob \< ow vs. ow \< ob ordering is controlled entirely by
$k$ — the smoothing parameter that determines how sensitive the learner
is to individual observations. At high $k$, the learner is insensitive
(like Model 1) and produces ob \< ow. At low $k$, idiosyncratic early
observations create within-class divergence before class structure is
detectable, and the learner produces ow \< ob.

**The key claim about** $k$: $k$ is like the inverse of a learning rate.
A large $k$ means each observation barely moves the estimate (low
effective learning rate). A small $k$ means each observation strongly
shapes the estimate (high effective learning rate). This is not an
analogy — it is a mathematical consequence of the formula.

## Mathematical formulation

$$\hat{P}_\text{verb}(t) = \frac{\text{count}(t \mid v, n) + k}{n + k \cdot V} \tag{2}$$

**Bayesian interpretation**: equation (2) is the posterior mean of a
Dirichlet($k$)–Multinomial conjugate model, where the prior places $k$
pseudo-observations on each of the $V$ token types. The prior is flat
(uniform over all tokens) — it carries no class-level information.

### The k-as-inverse-learning-rate proof

If we observe one more token $t$ after $n$ observations, the change in
the estimated probability of $t$ is:

$$\Delta\hat{P}(t) = \frac{\text{count}(t) + k + 1}{n + 1 + kV} - \frac{\text{count}(t) + k}{n + kV} \approx \frac{1 - \hat{P}(t)}{n + kV}$$

The magnitude of this change is inversely proportional to $n + kV$. At
large $k$, even a small $n$ gives a large denominator, and each
observation barely moves the estimate. This is exactly analogous to a
small learning rate in gradient descent.

```{r k_learning_rate}
# Demonstrate: how much does one observation change the estimate at different k?
V      <- 1000
n_obs  <- 10    # after 10 prior observations
t_prob <- 0.05  # current estimated probability of token t (for illustration)

k_values_demo <- c(0.001, 0.01, 0.1, 0.5)
cat("Change in P(t) per new observation of t, at n_obs =", n_obs, ":\n\n")
for (k in k_values_demo) {
  delta <- (1 - t_prob) / (n_obs + k * V)
  cat(sprintf("  k = %.3f:  denominator = %6.1f,  delta_P = %.6f\n",
              k, n_obs + k * V, delta))
}
cat("\nAt k = 0.001, one observation is", round(
  (1 - t_prob) / (n_obs + 0.001 * V) /
  ((1 - t_prob) / (n_obs + 0.500 * V)), 0),
  "x more impactful than at k = 0.5\n")
```

### Why low k produces ow \< ob

At low $k$, a verb's first few observations strongly spike its estimated
distribution around the specific tokens that happened to be observed.
Two same-class verbs that observed *different* idiosyncratic tokens will
look highly dissimilar — within-class DJS rises quickly. But this
idiosyncratic spiking is random: there is no reason why "give"'s
distance to "sell" should be systematically smaller than "give"'s
distance to "run". The ob criterion (which requires between \> within
*reliably across all verbs*) has no signal to detect yet. ob fires only
later, once sufficient data has accumulated for the shared class-level
tokens to dominate.

### Why high k produces ob \< ow

At high $k$, all distributions remain near-uniform for many
observations. The structural asymmetry (within-class token sharing \>
cross-class token sharing) means that, whenever DJS values are non-zero,
between-class DJS is slightly larger than within-class DJS. The relative
ob criterion detects this asymmetry even at tiny DJS values. The
absolute ow criterion must wait for within-class DJS to reach 0.01,
which takes much longer under heavy smoothing.

## Worked example: low k vs high k

```{r model2_comparison}
# Simulate a single run with low k and high k to illustrate the difference.
# Using the same token pool and true distributions for both.

set.seed(1)
n_cross_m2  <- round(0.3 * N_PREFERRED)
n_within_m2 <- round(0.3 * N_PREFERRED)
n_idio_m2   <- N_PREFERRED - n_cross_m2 - n_within_m2

cross_tok_m2    <- seq_len(n_cross_m2)
within_A_tok_m2 <- n_cross_m2 + seq_len(n_within_m2)
within_B_tok_m2 <- n_cross_m2 + n_within_m2 + seq_len(n_within_m2)
idio_pool_m2    <- (n_cross_m2 + 2L * n_within_m2 + 1L):VOCAB_SIZE_M1

mk2 <- function(cls) c(cross_tok_m2, cls,
                        sample(idio_pool_m2, n_idio_m2, replace = FALSE))
A_tok_m2 <- replicate(N_A_M1, mk2(within_A_tok_m2), simplify = FALSE)
B_tok_m2 <- replicate(N_B_M1, mk2(within_B_tok_m2), simplify = FALSE)

PA_true <- bld(A_tok_m2)
PB_true <- bld(B_tok_m2)

MAX_NOBS <- 2000L
N_OBS_GRID_M2 <- unique(round(exp(seq(log(1), log(MAX_NOBS), length.out = 15L))))

# Pre-sample draws
set.seed(42)
A_draws_m2 <- lapply(seq_len(N_A_M1), function(i)
  sample(VOCAB_SIZE_M1, MAX_NOBS, replace = TRUE, prob = PA_true[i,]))
B_draws_m2 <- lapply(seq_len(N_B_M1), function(i)
  sample(VOCAB_SIZE_M1, MAX_NOBS, replace = TRUE, prob = PB_true[i,]))

run_model2_simple <- function(add_k) {
  ob <- NA_integer_; ow <- NA_integer_; streak <- 0L; ow_st <- NA_integer_
  for (n in N_OBS_GRID_M2) {
    sm <- function(d) { ct <- tabulate(d[seq_len(n)], nbins = VOCAB_SIZE_M1);
                        (ct + add_k) / (n + add_k * VOCAB_SIZE_M1) }
    P_hat <- rbind(do.call(rbind, lapply(A_draws_m2, sm)),
                   do.call(rbind, lapply(B_draws_m2, sm)))
    D <- pairwise_jsd(P_hat)
    if (is.na(ob)) {
      ns <- sum(sapply(seq_len(N_A_M1), function(i) suppressWarnings(
        wilcox.test(D[i,(N_A_M1+1):(N_A_M1+N_B_M1)],
                    D[i,setdiff(seq_len(N_A_M1),i)],
                    alternative="greater", exact=FALSE)$p.value) < P_THRESH)) +
            sum(sapply(seq_len(N_B_M1), function(j) { i <- N_A_M1+j; suppressWarnings(
        wilcox.test(D[i,seq_len(N_A_M1)],
                    D[i,N_A_M1+setdiff(seq_len(N_B_M1),j)],
                    alternative="greater", exact=FALSE)$p.value) < P_THRESH }))
      if (ns / (N_A_M1 + N_B_M1) >= CLASS_FRAC) ob <- n
    }
    jA <- D[seq_len(N_A_M1), seq_len(N_A_M1)]
    jB <- D[(N_A_M1+1):(N_A_M1+N_B_M1), (N_A_M1+1):(N_A_M1+N_B_M1)]
    wm <- (mean(jA[upper.tri(jA)]) + mean(jB[upper.tri(jB)])) / 2
    if (is.na(ow)) {
      if (wm > DJS_THRESH) { if (is.na(ow_st)) ow_st <- n; streak <- streak + 1L;
                              if (streak >= SUSTAIN) ow <- ow_st
      } else { streak <- 0L; ow_st <- NA_integer_ }
    }
    if (!is.na(ob) && !is.na(ow)) break
  }
  c(ob = ob, ow = ow)
}

res_low_k  <- run_model2_simple(0.001)
res_high_k <- run_model2_simple(0.5)

cat("Low k (0.001):  ob_nobs =", res_low_k["ob"],
    "  ow_nobs =", res_low_k["ow"],
    " →", ifelse(res_low_k["ow"] < res_low_k["ob"], "ow < ob", "ob < ow"), "\n")
cat("High k (0.5):   ob_nobs =", res_high_k["ob"],
    "  ow_nobs =", res_high_k["ow"],
    " →", ifelse(res_high_k["ow"] < res_high_k["ob"], "ow < ob", "ob < ow"), "\n")
```

**Linking hypothesis confirmed**: the same model, same token pool, same
onset criteria, same random seed — different $k$ produces different
ob/ow orderings. This directly demonstrates that $k$ (learning
sensitivity) is the governing variable, not the type of learner.

------------------------------------------------------------------------

# Running the full grid searches

## Installation

``` r
install.packages(c("furrr", "progressr"))
```

No other packages are needed. Both scripts use only base R plus
these two packages.

## Running

From the directory containing the scripts:

``` r
source("scripts/zero-sensitivity-learner.R")      # ~10 minutes with 6 workers
source("scripts/variable-sensitivity-learner.R")  # ~45 minutes with 6 workers
```

Adjust `N_WORKERS` at the top of each script to match your machine
(`parallel::detectCores() - 1` is a safe default). On a Linux server
with 20+ cores, set `N_WORKERS <- 20`.

## Output format

Each script produces a CSV file:

| File | Rows | Key columns |
|------------------|------------------|------------------------------------|
| `grid_results_model1.csv` | 108 | `frac_ob_lt_ow`, `mean_ob`, `mean_ow`, `mean_log_ratio` |
| `grid_results_model2.csv` | 432 | `add_k`, `frac_ob_lt_ow`, `frac_ow_lt_ob` |

Each row is one parameter combination. `frac_ob_lt_ow` is the fraction
of seeds (out of N_SEEDS = 50) where ob fired before ow.
`mean_log_ratio` = mean log(ow/ob) for seeds where ob \< ow — positive
and large means ob fired much earlier than ow.

## Interpreting results

**Model 1**: expect `frac_ob_lt_ow` to be near 1.0 across almost all
parameter combinations. Exceptions at the weakest signal settings (small
item_overlap − class_overlap gap) are expected.

**Model 2**: expect a clean transition from `frac_ow_lt_ob ≈ 1.0` at low
`add_k` to `frac_ob_lt_ow ≈ 1.0` at high `add_k`. The transition should
occur somewhere around `add_k = 0.1–0.5`.

------------------------------------------------------------------------

# Summary of linking hypotheses

This table summarises the key claim-to-code links, so the reader can
evaluate them directly.

| Claim | Mathematical statement | Implemented as | Where to verify |
|------------------|------------------|------------------|------------------|
| Model 1 is the k→∞ limit of Model 2 | E[P̂(n)] = α(n)·P_verb + (1−α)·Uniform | Equation (1) matches equation (2) expected value | §3 proof; `k_learning_rate` chunk |
| k controls sensitivity to observations | ΔP̂(t) ≈ 1/(n+kV) per observation | Denominator of smooth_verb() in model2_addk.R | §4 proof; `k_learning_rate` chunk |
| ob is a relative test; ow is absolute | ob: Mann-Whitney; ow: threshold 0.01 | wilcox.test() vs DJS_THRESH constant | §2.5; onset criteria chunks |
| item_overlap \> class_overlap ensures class signal | DJS(same-class) \< DJS(cross-class) by construction | GRID filters class_overlap \< item_overlap | §2.1; token_pool_example chunk |

```{r session_info}
sessionInfo()
```
